The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-10-05 article.tex
Small tweaks

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Adoption of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015}, and
defining fingerprints for organic reactions~\cite{Segler2016}.

@@

The growth of open source software over the last 10 years [REF] is evidence
of the ability of communities of developers to develop systems and
processes that lead to high quality software systems for long term
use. The CDK is no different. The adoption of automatic build systems
and quality control methodologies such as unit testing, automated
source code validation, and peer review by fellow developers have
greatly improved the stability of the library, While it has slowed
development somewhat, it has allowed for cleaning up interdependencies
between modules of code, and importantly, has improved the scalability
of the development model.  This has resulted in significant new
functionality in core APIs while maintaining the quality of code
depending on those core APIs.

Examples of new features supported by the improved development model
include InChI functionality~\cite{Spjuth2013}, greatly improved ring
detection algorithms~\cite{May2014}, improvements to the core atom
type perception module that now covers a much more comprehensive set
of elements, charge states and radical species than previous versions,
a more comprehensive fingerprinting API and many speed and stability
improvements.

>>>>

The open source cheminformatics community has made significant steps
forward recently~\cite{OBoyle2011b} as evidenced by the growing number
of tools and underlying toolkits, along with the usage of these
software components in a variety of applications.
The Chemistry Development Kit (CDK) is one of the tools developed
under the aegis of the Blue Obelisk 
and previously documented versions have been widely adopted~\cite{Steinbeck2003,Steinbeck2006}.
Use of the CDK ranges from inclusion of CDK functionality in
wrapper platforms such as Cinfony~\cite{OBoyle2008}, incorporation
within the R environment (rcdk~\cite{Guha2007}), and as plugins for
Taverna~\cite{Truszkowski2011}, 
KNIME~\cite{Beisken2013}, Cytoscape (ChemViz2~\cite{ChemViz2}), and for
Microsoft Excel (LICSS~\cite{Lawson2012}).
In contrast to scenarios that have made CDK functionality available in
larger systems, a number of projects have employed the CDK as a
general cheminformatics toolkit. Examples include 
jCompoundMapper~\cite{Hinselmann2011}, ScaffoldHunter~\cite{wetzel2009interactive}, OMG~\cite{Peironcely2012},
PaDEL~\cite{yap2011padel}, ReactPRED~\cite{ReactPRED}, SMSD~\cite{Rahman2009,Rahman2014,Rahman2016},
WhichCyp~\cite{Rostkowski2013}, MetaPrint2D~\cite{Carlsson2010}, and the IUPHAR/BPS Guide to Pharmacology database~\cite{Southan2016}.
A number of such tools were initially developed using older versions
of the CDK and are updated to new releases as they are made
available. Examples include Bioclipse~\cite{spjuth2007bioclipse,
spjuth2009bioclipse} and
AMBIT~\cite{jeliazkova2011ambit,jeliazkova2011ambitsmarts,kochev2013ambit}. The
CDK has also played a role in a number of chemical studies, such as finding
the maximally bridging rings in chemical structures~\cite{Marth2015}, and
prediction of organic reactions~\cite{Segler2016}.

@@

The growth of open source software over the last 10 years is evidence           % add REF?
of the ability of communities of developers to develop systems and
processes that lead to high quality software systems for long term
use. The CDK is no different. The adoption of automatic build systems
and quality control methodologies such as unit testing, automated
source code validation, and peer review by fellow developers have
greatly improved the stability of the library. While it has slowed
development somewhat, it has allowed for cleaning up interdependencies
between modules of functionality, and importantly, has improved the scalability
of the development model.  This has resulted in significant new
functionality in core APIs while maintaining the quality of code
depending on those core APIs.

Examples of new features supported by the improved development model
include InChI functionality~\cite{Spjuth2013}, greatly improved ring
detection algorithms~\cite{May2014}, improvements to the core atom
type perception module that now covers a much more comprehensive set
of elements, charge states and radical species than previous versions,
a more comprehensive fingerprinting API, new depiction functionality,
and many speed and stability improvements.
